---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r, results="hide"}
df <- read.csv("activity.csv")
```

Above line reads and creates a data frmae for processing the csv file

## What is mean total number of steps taken per day?

A. Calculating the total number of steps taken per day

1) Total number of steps taken per day is calculated by using tapply calling sum of steps with date as factor.
2) tapply provides the result as an array.
3) This is converted to data frame using as.date.frame function

```{r, results="hide"}
stepsPerDay <- tapply(df$steps, df$date, FUN=sum, na.rm=TRUE)
res <- as.data.frame(stepsPerDay)
```

B. Histogram of the total number of steps taken each day

```{r}
hist(res$stepsPerDay,main="Histogram of steps taken each day",xlab="Number of Steps")
```

C. Calculating and reporting the mean and median of the total number of steps taken per day

```{r, results="hide"}
meansteps <- mean(res$stepsPerDay)
medsteps <- median(res$stepsPerDay)
```

Mean of total number of steps taken per day is `r meansteps`.
Median of total number of steps taken per day is `r medsteps`.

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l" ) of the 5 minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
av <- tapply(df$steps, df$interval, FUN=mean, na.rm=TRUE)
avs <- as.data.frame(av)
avs$interval <- rownames(av)
plot(avs$interval, avs$av, type="l",xlab="Interval", ylab="Number of Steps")
```

2. Which 5 minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, results="hide"}
maxav <- max(avs$av, na.rm=TRUE)
maxdf <- avs[avs$av==maxav,]
maxintervl <- maxdf[1,2]
```

Interval with max number of steps is `r maxintervl`

## Imputing missing values

1. Total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
sum(is.na(df$steps))
```

2. Approach for handling NAs

When I executed the below code, it was observed that 288 intervals are reported for each date.

```{r, results="hide"}
library("plyr")
count(df$date)
```

When I executed the below code, it was observed that 8 dates contained all 288 intervals filled with NAs.

```{r}
count(df$date[is.na(df$steps)])
```

Since 8 dates did not have value, mean will vary if denominator decreases. Hence decided to check how the values differ if these dates were removed from the data set for analysis.

3. Created a new dataset by removing NA value rows from the table

```{r, results="hide"}
cdf <- na.omit(df)
```

4. Histogram of the total number of steps taken each day

```{r}
uptdsteps <- tapply(cdf$steps, cdf$date, FUN=sum, na.rm=TRUE)
reslt <- as.data.frame(uptdsteps)
hist(reslt$uptdsteps,main="Histogram of steps taken each day in revised data set",xlab="Number of Steps")
```

while comparing the initial one, total number of steps seems to be reduced.

## Are there differences in activity patterns between weekdays and weekends?

1. Created a new factor variable dayType in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r, results="hide"}
# Using weekdays built-in function find the day of the week
# if it starts with S i.e. Saturday or Sunday, categorize as weekend, else weekday

finddaytype <- function(x) {
  ifelse (grepl("^S",weekdays(as.Date(x,'%Y-%m-%d'))) == 1,"weekend","weekday")
}

# Using lapply, execute this function for the list of all dates i.e. date column in the data frame
df$dayType <- lapply(df$date,finddaytype)

```


2. A panel plot containing a time series plot (i.e. type = "l" ) of the 5 minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
wkendf <- df[df$dayType=="weekend",]
wkdayf <- df[df$dayType=="weekday",]

avwkend <- tapply(wkendf$steps, wkendf$interval, FUN=mean, na.rm=TRUE)

avwkday <- tapply(wkdayf$steps, wkdayf$interval, FUN=mean, na.rm=TRUE)

avwkends <- as.data.frame(avwkend)
avwkdays <- as.data.frame(avwkday)

avwkends$interval <- rownames(avwkend)
avwkdays$interval <- rownames(avwkday)

#par(mfrow = c(2, 1), width = 480, height = 1000)

plot(avwkends$interval, avwkends$av, type="l",main="Weekend", xlab="Interval", ylab="Number of Steps")
plot(avwkdays$interval, avwkdays$av, type="l",main="Weekday",xlab="Interval", ylab="Number of Steps")


```

__Conclusion :- __ Graphs on weekend vs weekday clearly shows that there is more activity during weekend.